Mining Interaction Log Data in a Creative Arts Practice using Macro Gestures and Transition Matrices

### 2015-09-21 flux-entropy-paper/flux-entropy-paper.tex
lots of introduction edits

Modern computing devices and interfaces continue to open up rich,
open-ended creative opportunities in the new media arts. Understanding
and evaluating interfaces in these contexts presents many
challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}. In the context of
live performance, interface evaluation, using, for example,
post-performance questionnaires, can omit consideration of the dynamic
needs of artists in the moment of the performance and in the wild of
the performance context. In contrast to traditional art forms, logged
interaction data from a new media arts performance can provide an
enduring record of that performance for post-hoc quantitative
investigation as well as for curation and
preservation~\cite{England:2014ys}. Over time, the accumulation of
new-media interaction log data might allow data-mining techniques to
be deployed to potentially describe universal features of such
interaction.

@@

It is a feature of creative expression in the arts that highly skilled
practitioners exploit simple tools to produce works of beauty whether
it be static paintings or dynamic performance. In previous
work~\cite{Martin:2014cr} researchers have shown that trained
percussionists can exploit a very simple iPad interface to discover
new categories of musical gestures. In that paper, iPad apps were
studied where tapping was mapped to produce produced short sounds,
while swiping produced long sounds with the volume proportional to the
velocity of the moving touch point. It turned out that trained
percussionists working together in an ensemble were able to discover
and repeatedly make use of five macro gestures made up of taps and
swipes: fast taps, slow taps, random taps, short swipes and swirls.
This process of discovery of interesting musical gestures from a
simple interface is analogous to what percussionists do with acoustic
instruments -- drums and bowls and bells are often tapped, stroked and
hit in novel ways to communicate meaning and emotion through
music~\cite{Schick:2006fk}.

In this note we describe a log data archive that has been accumulated
to support a long-term practice-led research program into simple iPad
interfaces for group music-making similar to those described in [ref:
charles at CHI] . Over a period of more than two years (April
2013--June 2015), we collected data from $....$ group music-making
sessions where the participants used custom-built, touch-screen apps
on Apple iPad devices. Although several different apps have been
included in our corpus,
% TODO: should there be a citation here?
they all shared the simple mode of interaction where the majority of
the screen could be used for free-form touch interactions; tapping
produced short sounds, while swiping or swirling produced long sounds
with the volume proportional to the velocity of the moving touch
point. Apart from this commonality, our apps featured different
palettes of available sounds and modes of synthesis, different
arrangements of sounds on the iPad screens, different kinds of visual
display and feedback, different kinds of networked interactions
(between the iPads in an ensemble and with a central server), as well
as a variety of special features for each app.

@@

% TODO: NOTE: FORGET ABOUT THE DEMONSTRATIONS AND THE DATA COLLECTION
% PARTS OF THE CORPUS. THEY JUST MAKE THE STORY VAGUE. YOU DON'T
% INCLUDE THEM IN YOUR ANALYSIS AT ALL.

% TODO: NOTE: NEED TO MODIFY ALL OF THESE GRAPHS, TABLES AND FIGURES.

\begin{table}
%\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 150       &         &        &  \\
Record size & 306MB     &         &        &   \\
Length      & 20H34M22S & 1M0S    & 7M29S  & 26M0S\\
Participants& 524       & 1       & 4      & 9    \\
Flux        &           & 0.02239 & 0.1445 & 0.3246\\
\hline
\end{tabular}
\caption{
  Representative data about the corpus of musical touch-screen
  interaction sessions used in this study. These sessions were
  rehearsals, performances, studies, and demonstration sessions and 
  participants improvised or performed compositions. Each record
  consisted of touch-data recorded in CSV format and was later analysed
  with a gestural classification system. Some participants were
  present in multiple sessions, however this data was not tracked in
  the corpus.\label{corpus-table}}
\end{table}

@@

A useful way of modelling user interfaces is to consider their
configurations as state vectors and to deploy the methods of matrix
algebra to describe transitions between these states during an
interaction sequence\cite{Thimbleby:2001kq, Thimbleby:2004fj}. Data
obtained from usability analysis of quite complex systems (such as 
resource planning software for security applications~\cite{Kannampallil:2007fp})

have been undertaken by coding sequences of fine-detail human actions
as states and then to model data from usability walkthroughs as
sequences of state transitions. The transition probably matrix for
transitions between states can then be derived from such user data and
conclusions drawn about the nature of interaction. In the creative
arts, editing commands at a computer interface for the ``live coding
of computer music'' have also been coded coded as states and analysed
using transition matrices, in both the textual and musical dimensions,
in order to arrive at conclusions regarding artistic
style\cite{Swift:2014tya}.

In our corpus of iPad musical performances, rehearsals and studies,
the macro gestures identified by Martin et al~\cite{Martin:2014cr}
provide a basis to identify and code macro gestures. Each performer in
each muscial piece can then be thought of as following a ``score'' of
state transitions and an overall transition matrix for the entire
ensemble can be derived by averaging the individual transition
matrices.

@@

Previous work showed that the raw touch data obtained from interaction
with simple iPad interfaces similar to ours could be segmented using a
Random Forest\cite{Breiman:2001kx} classifier\footnote{The classifier
  used was from Python's Scikit-learn package\cite{scikit-learn}.}.
The classifier was applied to 5-second windows of the touch-data to
provide a chain of gesture-states for each participant in each
session. These chains can be used to create $9 \times 9$ (the
classifier recognises 9~distinct gestures) transition matrices for
individual performers and the whole ensemble in each session.

@@

In the rest of this paper we will use these matrix measures to answer
two questions. First, do the transition matrix measures allow us to
distinguish between different performance contexts, types and
instrumental setups? Secondly, do these performance measure allow us
further insight into the corpus than the metadata can provide?
% TODO I think this second question is too broad?

@@

\begin{figure} \centering
  \includegraphics[width=\linewidth]{figures/context-flux-entropy-boxplot}
  \caption{Distributions of flux and entropy values by performance
context. Study and rehearsal sessions had the highest flux for
improvisations which fell during performances. Compositions had lower
flux in rehearsals than performances. Demonstration sessions had much
lower entropy indicating data unevenly spread in the transition
matrices. This could be because untrained participants did not explore
the full gestural potential of the instruments.
\label{fig:flux-entropy-boxplot}}
\end{figure}

>>>>

Modern computing devices and interfaces continue to open up open-ended
creative opportunities for new artistic experiences. Understanding and
evaluating interfaces in these contexts presents many
challenges~\cite{Resnick:2005yu,Shneiderman:2007qv}.

In the context of live performance, interface evaluation, using, for
example, post-performance questionnaires, can omit consideration of
the dynamic needs of artists in the moment of the performance and in
the wild of the performance context.

In contrast to traditional art forms, logged interaction data from
interaction with a computer-based artistic interface can provide an
enduring record of that performance for post-hoc quantitative
investigation as well as for curation and
preservation~\cite{England:2014ys}.

Over time, the accumulation of
interaction log data might allow data-mining techniques to be deployed
to potentially describe universal features of such interaction.

@@

It is a feature of creative expression in the arts that highly skilled
practitioners exploit simple tools to produce works of beauty whether
it be static paintings or dynamic performance.

In previous work~\cite{Martin:2014cr} researchers have shown that
trained percussionists can exploit a very simple iPad interface to
discover new categories of musical gestures.

In that paper, iPad apps
were studied where tapping was mapped to produce produced short
sounds, while swiping produced long sounds with the volume
proportional to the velocity of the moving touch point.

It turned out that trained percussionists working together in an
ensemble were able to discover and repeatedly make use of five macro
gestures made up of taps and swipes: fast taps, slow taps, random
taps, short swipes and swirls. This process of discovery of
interesting musical gestures from a simple interface is analogous to
what percussionists do with acoustic instruments -- drums and bowls
and bells are often tapped, stroked and hit in novel ways to
communicate meaning and emotion through music~\cite{Schick:2006fk}.

In this note we describe a log data archive that has been accumulated
to support a long-term practice-led research program into simple iPad
interfaces for group music-making similar to those described in Martin
et al~\cite{Martin:2014cr}. Over a period of more than two years
(April 2013--June 2015), we collected data from group music-making
sessions where the participants used custom-built, touch-screen apps
on Apple iPad devices. 
Although several different apps have been
included in our corpus, they all shared the simple mode of interaction
where the majority of the screen could be used for free-form touch
interactions; tapping produced short sounds, while swiping or swirling
produced long sounds with the volume proportional to the velocity of
the moving touch point. Apart from this commonality, our apps featured
different palettes of available sounds and modes of synthesis,
different arrangements of sounds on the iPad screens, different kinds
of visual display and feedback, different kinds of networked
interactions (between the iPads in an ensemble and with a central
server), as well as a variety of special features for each app.

@@

% TODO fix this table to remove descriptive data about the
% demonstration sessions.
\begin{table}
\centering
\begin{tabular}{l|llll}
\hline
            & Total & Min  & Median   & Max     \\ 
\hline
N           & 150       &         &        &  \\
%Record size & 306MB     &         &        &   \\
Length      & 20H34M22S & 1M0S    & 7M29S  & 26M0S\\
Participants& 524       & 1       & 4      & 9    \\
Flux        &           & 0.02239 & 0.1445 & 0.3246\\
%Entropy     &           &  &  & \\          
\hline
\end{tabular}
\caption{
  Representative data about the corpus of musical touch-screen
  interaction sessions used in this study. These sessions were
  rehearsals, performances, studies, and demonstration sessions and 
  participants improvised or performed compositions. Each record
  consisted of touch-data recorded in CSV format and was later analysed
  with a gestural classification system. Some participants were
  present in multiple sessions, however this data was not tracked in
  the corpus.\label{corpus-table}}
\end{table}

@@

A useful way of modelling user interfaces is to consider their
configurations as state vectors and to deploy the methods of matrix
algebra to describe transitions between these states during an
interaction sequence~\cite{Thimbleby:2001kq, Thimbleby:2004fj}.
An opposite approach is to record users' interactions with an
interface, manually coding interaction states. 
This method has been applied to usability analysis of resource
planning software for security applications~\cite{Kannampallil:2007fp}
where the transition probability matrix of these state sequences
was used to draw conclusions about the nature of this interactions.
In the creative arts, editing commands at a computer interface for the
``live coding'' of computer music has also been coded as states and
analysed using transition matrices~\cite{Swift:2014tya}, in both the
textual and musical dimensions, in order to arrive at conclusions
regarding artistic style.

In our corpus of iPad musical performances, rehearsals and studies, we
have used the continuous touch-screen gestures identified by Martin et
al~\cite{Martin:2014cr} as a basis for a musical state-space.
Each performer in each musical piece can then be thought of as
producing a ``score'' of state transitions and an overall transition
matrix for the entire ensemble can be derived by averaging the
individual transition matrices.

@@

Previous work~\cite{Martin:2015jk} showed that the raw touch data
obtained from interaction with simple iPad interfaces similar to ours
could be segmented using a Random Forest\cite{Breiman:2001kx}
classifier\footnote{The classifier used was from Python's Scikit-learn
  package\cite{scikit-learn}.}. The classifier was applied to 5-second
windows of the touch-data to provide a chain of gesture-states for
each participant in each session. These chains can be used to create
$9 \times 9$ (the classifier recognises 9~distinct gestures)
transition matrices for individual performers and the whole ensemble
in each session.

@@

In the rest of this paper we will use these matrix measures to answer
two questions. First, do the transition matrix measures allow us to
distinguish between different performance contexts, types and
instrumental setups? Secondly, do these measures allow us further
insight into the internal structure of performances?

%  Secondly, do these performance measure allow us
% further insight into the corpus than the metadata can provide?
% % TODO I think this second question is too broad?

@@

\begin{figure} \centering
  \includegraphics[width=\linewidth]{figures/context-flux-entropy-boxplot}
  \caption{Distributions of flux and entropy values by performance
context. Study and rehearsal sessions had the highest flux for
improvisations. Compositions had lower
flux in rehearsals than performances.
\label{fig:flux-entropy-boxplot}}
\end{figure}
